Modal Verification Patterns for Systems

### 2025-06-16 paper.tex
+

\paragraph{Virtualization of Disk-Blocks in Filesystems}

\paragraph{Consistency of Disk Layout} \todo{}
\paragraph{Memory Reclamation in Client Code}\todo{iso:speak about Amal's stuff}
\paragraph{Consistency of Reclaimed Memory}\todo{iso: speak about the operator for which the hybrid logic stuff is usefull }

>>>>

\paragraph{Virtualization of Disk-Blocks in Filesystems}
\textsf{ModFS} is a copy-on-write (\textsf{CoW}) filesystem, which as described in the introduction performs updates by copying modified data to \emph{new and unused} locations on disk rather than updating in place, ensuring that in the absence of physical storage corruption, there is \emph{always} a fully consistent filesystem on disk, removing the need for recovery from journals in the case of a crash, which can be very expensive for large volumes.
%\textsf{ModFS} is implemented in \textsc{Go}~\cite{donovan2015go}, using the FUSE~\cite{fuse} interface to run a filesystem in usermode. \textsf{ModFS}'s code is translated to \textsf{Roqc} using the \textsf{Goose}~\cite{chajed2019verifying,chajed2021gojournal} tool, which translates a significant subset of \textsf{Go} to \textsf{Roqc} to permit formal verification of \textsc{Go} code using \textsf{Iris}~\cite{jung2018iris}. \textsf{Goose} generates code in an extension of \textsf{Iris}'s \textsf{HeapLang} with libraries for disk manipulation and \textsc{Go} primitive types (e.g., maps) as shown in Figure \ref{fig:modfslayout}.

%On the left part of Figure \ref{fig:modfslayout}, we see the shaded region -- \textsf{ModFs} storage layer-- under the yellow layer --- the \textsf{ModFs} system call layer. 
%We have verified the parts of the shaded storage layer involved in creating a new on-disk snapshot, up to an axiomatized specification of of the \textsf{Bcache} component that implements a simple in-memory cache of disk blocks.
%This primarily involves the \textbf{write\_amplification} function on btree nodes (from the \textsf{Btree} module), and its associated dependencies, including the \textbf{split}, \textbf{put}, and \textbf{del}(ete) operations on nodes, and the \textbf{write} serialization function.
%This also includes snapshot operations including \textbf{snap\_initialize} and \textbf{snap\_alloc} functions (among others) in the \textsf{Snapshot} module, and freelist operations (\textbf{free} and \textbf{allocate}) from \textsf{Freelist}, as well as the supporting functions from the \textsf{Chunk} (roughly a structured disk block abstraction), \textsf{PageMap} (raw disk block abstraction), \textsf{and Lock} (thread synchronization via a global lock map) modules.
%\textsf{Storage} (a data structure of runtime pointers\colinsays{doesn't really explain it, I don't know what this component is/does}), \textsf{BtrItr} (an iterator library), \textsf{inode}, and \textsf{Btree} have not yet been fully verified, though the components involved in write amplification have been verified. The filesystem does run and standard filesystem functionality (core file and directory operations) work.
%This paper focuses on the verification of write amplification, which is verified up to a small number of axioms for \textsf{Go} data structures\colinsays{update w/ very short summary of remaining admits} and the unverified \textsf{Bcache} specification.
%\colinsaysinl{How many lines of code are covered by the verification? (excluding the bcache) This might help assuage reviewer concerns that we're not done, by making the scope of this project more explicit so ``why didn't you finish'' has a self-evident answer of ``because they already did a ton of work''}

%\textsf{ModFs} is a \textsf{CoW} file system where the mapping of the logical locations, e.g., file pages indexed by file offsets, to disk locations by employing a \textsf{CoW} tree \cite{Astrahan1976SystemR,shadowclone} structure. In Figure \ref{fig:cow_index_tree}, we see that accessing and updating the $\textsf{page}_n$ of file \textsf{"foo"} referred by the file offset $\textsf{offset}$ 
%\[ \textsf{write}(\textit{foo},\textsf{offset})\]
%requires locating the corresponding physical disk block of the in-memory $\textsf{page}_n$ whose mapping is establised the \textit{copy-on-write} index tree (in the middle layer). Like Btrfs ~\cite{}, \textsf{ModFs} indexes all filesystem objects (e.g. files and directories) as a nested trees, and to map a logical page to a physical page, the file system first locates the root of the nested index-tree belonging to \textsf{"foo"}.
Semantically, in \textsf{ModFS} (as in other \textsf{CoW} filesystems~\cite{young1992episode,bonwick2003zettabyte,rodeh2013btrfs,Rosenblum1992LFS,hitz1994file}) any update to file data or metadata is realized on  unused disk locations by treating the tree as a persistent data structure. 
As an example, Figure \ref{fig:cow_index} shows that an update to the file $\textsf{"foo"}$ at $\textsf{"offset"}$ ends up with locating the related disk chunk (a set of blocks) in the gray region, copying it to newly allocated disk chunks (dotted chunks on the right side of the bold long-dash-point line of the middle layer withing dotted section), and performing the update on these newly-allocated disk blocks. This requires further updates to those metadata blocks which previously referred to the old block locations to instead reference the newly-allocated blocks. This must be repeated until the allocation/update of the new chunks percolates up to the root (including the root), through a process called \textsf{\textit{write-amplification}}.

\begin{figure}
    \includegraphics[width=0.75\columnwidth]{cow_index.png}
    \caption{A Filesystem called \textsf{Modfs} Indexing its Physical Pages for File Pages with CoW (copy-on-write) Tree}
    \label{fig:cowindex}
\end{figure}

After this process, traversing the filesystem from either the old or new root yields consistent views of the filesystem at different points in time, each commonly referred to as a \emph{snapshot}.
These snapshots overlap on the unmodified portions of the filesystem (the left-most grid section), with the reclaimed the chunks (the middle gray section) becoming available for reuse once the new snapshot becomes the official current state of the filesystem.\footnote{Many of these filesystems permit keeping archived snapshots of the filesystem state, e.g., a snapshot of the system every hour up to a certain number of days, to support user data recovery. In these cases only disk regions that are not still in use by an archived snapshot are reclaimed.}. This permits the filesystem to atomically switch between consistent on-disk states, by updating the oldest so-called \emph{super} chunk at a fixed location on disk.\footnote{Filesystems must take care to ensure that possibly-flaky hardware truly does perform an atomic update.} Super chunks, as we see in Figure \ref{fig:cow_index} as ($\textsf{Super}_0$ and $\textsf{Super}_1$), are the first two blocks of the disk dedicated to keep the state of the file system which is composed of some meta-data on the filesystem, and most importantly to the root chunk of the file system index-tree. Therefore, in case of a crash during the update to the \textsf{file "foo"}, we can reboot the file system from the previous persisted root, \textsf{root chunk}, via reading the $\textsf{Super}_0$. This means the whole disk consistency checks are unnecessary when recovering from a crash~\cite{bonwick2003zettabyte,hitz1994file}; recovery is immediate, as opposed to requiring hours or days~\cite{henson2006reducing,agrawal2009speedy} (or longer~\cite{henson2006chunkfs}) for recovering a large volume’s consistency from logs. 
An important additional subtlety with this kind of filesystem is \emph{when} blocks in the reclaimed region can actually be reused: only \emph{after} the snapshot that made those blocks ``free'' has been completely written to disk. Allowing reuse before this could result in in-place updates: part of the new snapshot could attempt to store data in space used by the prior snapshot, allowing inconsistent on-disk structures. This means the filesystem must carefully track not only what blocks are reclaimed, but \emph{when} they are safe to reuse, making the \textsf{Freelist} component of \textsf{Modfs} and other \textsf{Roqc} filesystems non-trivial.
With their intricate semantics and critical place in modern systems, formal verification of functional correctness for storage systems (including filesystems and other kinds of storage) have drawn much attention~\cite{ntzik-gardner-15,ntzik2018concurrent,chen-crash-16}\todo{lots more cites}. Unlike our \textsc{CoW} filesystem \textsf{ModFs} which allocates new chunks and rewrites all the nodes from leaf to root on the write-amplification path, some filesystems allows in-place updates to handle the heavy workloads more efficiently in principle. Unlike \textsc{CoW} filesystems, filesystems with in-place updates need to implement some form of \emph{journalling} to ensure that a crash partway through an in-place update can be rolled back (using an undo journal that records the original contents before modification) or completed (using a write-ahead journal that first writes new data in unused space before \emph{also} updating in place).\footnote{\textsf{CoW} filesystems can be viewed as a twist on write-ahead journals that uses the journal as the actual data once written to a fresh location, thereby reclaiming the original location. This results in a performance and complexity trade-off: journalling writes twice for every updated block and requires recovery logic to roll back or re-execute the log, while \textsf{CoW} techniques write each update only once and require slightly more intricate allocation strategies to compact fragmentation.}
\paragraph{In-Place Updates in Filesystems}In-place updating filesystems with journalling components have been specified and verified with an extended Hoare Logic to specify not only the pre/post conditions of the operations successful operations in the filesystem, but also the crash condition~\cite{chen-crash-16,chen2017verifying}, with later work extending this for concurrent filesystem implementations~\cite{chajed2021gojournal}. \textsf{Perennial}~\cite{chajed2019verifying} extends the weakest precondition of the \textsf{Iris} separation logic~\cite{jung2018iris,krebbers2017essence} to allow asserting facts about disk state at crashes (the \emph{crash condition}) and apply this extension to a journalling filesystem \textsf{GoJournal}~\cite{chajed2021gojournal}. Unlike \textsf{GoJournal}, \textsf{ModFs} relies on \textsf{CoW} semantics for consistency, eliminating the need for recovery logic.
Essentially, successfully writing the new superblock for a new snapshot is analagous to completing a valid journal entry, except it \emph{also} has the effect of updating the live filesystem data due to making the ``journal entries'' valid filesystem data and metadata blocks (an idea that goes back to early work on log-structured filesystems~\cite{Rosenblum1992LFS}).
In fact, we can think of the atomicity of the recovery from the crash state conceptually equal to the single block disk write persisting the change from $\textsf{Super}_0$ to $\textsf{Super}_1$ in Figure \ref{fig:cow_index_tree_update}, therefore, considering \textsc{CoW} semantics, we do not need to consider the specification and proof of crash condition separately; instead a global never-violated invariant (not even temporarily broken) of the filesystem ensures the on-disk state is always consistent.
\paragraph{Consistency of Disk Layout} \todo{}

\paragraph{Memory Reclamation in Client Code}\todo{iso:speak about Amal's stuff}
\paragraph{Consistency of Reclaimed Memory}\todo{iso: speak about the operator for which the hybrid logic stuff is usefull }
